---
title: "BARGAINING OVER BEAUTY"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: true
    toc_depth: 1
    #code_folding: hide
    abstract: This paper argues that concerns over opportunism affected the content and structure of Renaissance art contracts. Building on insights from the economic analysis of contracts, we first show that the relationship between buyer---the patron---and seller---the painter---in Renaissance Italy was exposed to the threat of opportunism and other sources of conflict. We then test the influence of these threats to the contracting process for paintings and test these against a novel data set on the content and structure of ninety-one documents corresponding to as many commissions. Our results provide evidence that concerns over opportunism had a systematic effect on the trading parties' choice of how much and what to include in the contract governing their exchange.
bibliography: references.bib
biblio-style: apa
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

# Introduction

Contracts are a key market institution. They facilitate trade by providing trading partners with essential information about their respective rights and responsibilities. Just as important is their mitigation of opportunism by economic actors. Trading parties who feel they have been cheated can leverage the existence of a contract in courts to force the cheater to compensate them for any economic damages. Unsurprisingly, economists have produced a large literature on contracts---their structure and content, their impact on economic activities, and more. One important insight from this literature is that, although formal contracts---written documents that use formulaic language to specify each party's claims and obligations and are enforceable by a third party, usually the government---benefit all those involved, they are also costly to draft and enforce. Thus, we seldom, if ever, observe "complete" contracts in real-world markets.

This paper builds on the contributions from the literature on the economics of contracts to study the fundamental institution that governed the market for paintings in Renaissance Italy: The commission contract. On the demand side of this market were the patrons---those individuals or corporate entities willing to disburse significant sums of money to decorate public spaces and signal their magnificence to the rest of society [@nelson2014patron]. On the supply side were master painters and their workshops of skilled artisans. When the two sides came together---i.e., when the patron awarded a commission to a master painter---they often wrote a document, aided by a legal professional or a notary. This document was a legally binding contract listing both parties' obligations to each other. Several features of Renaissance commission contracts are of interests to economists. First, their content strongly indicates that both patrons and painters worried about the possibility each other's opportunism. Second, their content and structure varied significantly from commission contract to commission contract. Some were quite long, referenced several categories of obligations explicitly, and contained detailed descriptions of the final product. Some were quite short, referenced only a few duties of either party, and made no effort at all to describe the subject of the painting.

A cursory understanding of the workings of Renaissance art markets explains why patrons were concerned with opportunism. Paintings from this period were often produced with expensive materials like gold and lapis lazuli, which painters may replace with much cheaper options and pocket the difference. Painters whose work was popular with many potential patrons would also often be tempted to take one too much work at once, leaving them with the choice between subcontracting much of it to their (less talented) assistants or failing to execute commissions by the agreed upon date. Neither option would have been particularly popular with patrons. Nor where patrons immune from the lure of opportunism. A painter may rightfully fear that the patron may renege on their promise to pay, perhaps by claiming to be unsatisfied with the final painting.

To identify and measure the effect of the threat of opportunism on contracting practices in Renaissance art markets, we build a novel data set using commission contracts from this period. We tracked down documents corresponding to commissions for paintings from Renaissance Italy from a wide array of secondary sources, most importantly the list of commission contracts in @omalley2005business and those in the *Italian Renaissance Document Site* from the University of Kent. We then analyzed each document to extract information on the terms and conditions agreed upon by the trading partners. In total, out data set consists of ninety-one documents, each corresponding to a unique painting commission. For each document, we created ten binary variables corresponding to ten categories of terms and conditions found in Renaissance art contracts and used this information to generate an index variable to measure contractual completeness.

We combine these data with commission-, patron-, and painter-specific information, including whether the commission was awarded by a secular or religious institution, the age, years active as painter, and number of paintings completed by the latter at the signing of the contract. We use the resulting data set to test a variety of hypotheses on how the threat of opportunism shaped the content and structure of Renaissance art contracts. Our results show a large, robust, and statistically significant effect of the value of a commission---the total compensation promised by the patron to the painter---and the degree of contractual completeness of the commission document---whether we measure it by the length of the same document or by our index of contractual completeness. We find also suggestive evidence that a painter's contemporaneous reputation---measured alternatively in terms of his age, the length of his career up to that point, and the number of commissions completed until the year before---is negatively correlated with contractual completeness.

Our analysis relate to three strands of economic literature. First, we contribute to the study of the economics of contracts. More specifically, our argument builds on the literature on the role of transaction costs on institutional choice [@williamson1979transaction, @barzel1997economic]. This literature emphasizes the threat of opportunism as a determinant of trading parties' decision of how to structure their interactions [@klein1978vertical]. Another essential intuition associated with this body of work is that the governance of exchange is costly: Contracts can never be complete---in the sense that they cannot make provisions for all possible future circumstances---and their enforcement is not free. Real-world markets function thanks to a mix of formal and informal contracts, always short of complete and enforced by a combination of public institutions---like government-run court---, private institutions---such as arbitrators---, and market forces---like vertical integration and reputation [@macleod2007reputations]. We contribute to this literature by providing systematic evidence that concerns for opportunistic behavior affect the content and structure of the contract governing the same transaction. This is consistent with the argument in @al1995incomplete that the characteristics of trading partners, legal environment, and the commodity exchanged will determine the equilibrium degree of contractual completeness of a given transaction. Moreover, our results support claims about the substitutability of reputation and formal contractual enforcement [@klein1981role, @barzel1997economic]

The second literature our argument relates to is the economics of contracts in a variety of historical settings. @umbeck1977theory is an early contribution to this literature. Building on the model of contract choice developed by @cheung1968private, Umbeck argues that gold miners during the California gold rush of the 1840s arranged their interactions in ways predicted by the theory. For instance, he finds that gold miners abandoned the sharing model was the costs of enforcement increased significantly due to the rapid growth in the mining population. In a similar vein, @lueck1994common shows that the institutions governing access to common pool resources by pre-modern societies followed a similar logic.^[Other applications of the theory of contracts to historical institutions include @dari2013slavery on slavery in the antebellum American South, @leeson2007arrgh on eighteenth century pirates, @silverman2017asset on maritime trade in the eighteenth century, and @ellickson1989hypothesis on the whaling industry in nineteenth century New England.] A distinct body of work, although also related to the economic theory of contracting, is that which emphasizes the role of private institutions and market forces in the enforcement of trade. For instance, @milgrom1990role argue that the commercial revolution of the Late Middle Ages had been made possible by private courts that relied on a body of judicial rules known as "law merchant" and whose decisions were enforced by ostracism from merchant-run trade fairs. Another influential contribution is @greif1989reputation, which shows that medieval Jewish merchants were able to succeed in the business long distance trade in the absence of formal enforcement. Instead, these traders relied on informal reputational mechanisms that required the communication of information about each trader's past performance between members of their "coalition." The threat of seeing future business opportunities damaged by a loss of reputation was enough to guarantee performance in most circumstances.^[Reputation-based mechanisms can be found in a variety of historical contexts. For instance, @leeson2007trading argues that they facilitated trade between Europeans and Sub-Saharan Africans in the pre-colonial period. According to @allen2009theory, reputational considerations shaped the inner working of the pre-modern British aristocracy and assured their performance as members of the royal aristocracy.] We contribute to this body of work by extending the same theoretical framework to Renaissance art markets in the Italian peninsula. This extension is of particular interest due to the nature of the commodity and the institutional context of the transaction. Renaissance paintings were complex commodities, their value in large part a function of characteristics---like the quality of their design and their composition---that are hard to contract upon ex-ante. These characteristics would make third-party enforcement challenging even in the presence of a well-functioning legal system like those found in advanced market economies. Yet, a market for Renaissance art arose and thrived in Renaissance Italy, before the development of modern public institutions to support private transactions.^[However, we must note that Renaissance Italy, especially its most successful city-republics, enjoyed more effective public institutions than much if not most of the rest of Western Europe at the time [@jones1997italian].]

Finally, we contribute to the economics of Renaissance art markets. @etro2013market and @etro2018economics are recent contributions to this field. They find that the evidence on pricing in the market for paintings in Renaissance and Early Modern Italy is consistent with economists' standard understanding of market dynamics. For example, they find that prices converged toward one price once painters' reputation and the physical characteristics of the paintings were accounted for. An earlier contribution by @owen1977florentine relies on monopsony theory to argue that stylistic differences between the Florentine and Sienese schools of painting---the two major schools in central Italy throughout the early Renaissance---were due to a difference in sensibilities between the elites of the two Tuscan cities. In Florence, humanism and realism were dominant while traditionalist and religious aesthetic values prevailed among the Sienese establishment. Perhaps the work that overlaps most significantly with ours is that of @omalley2005business. An art historian, O'Malley studied over one hundred documents from the Renaissance period to shed light on the process of bargaining between patrons and masters over the price, content, and style of Renaissance paintings.^[Also related to our contribution is @omalley2013painting, which explores the issue of artists' reputation in facilitating transactions with patrons.] Our paper is the first in this literature to a) make explicit use of the insights from the economics of contracts and b) rely on systematic empirical analysis to study the structure and content of commission contracts during this period.

# Historical background

## The market for paintings in Renaissance Italy

The Renaissance period in Italian art history starts with the last few decades of the thirteenth century, when Giotto revived the forgotten art of painting, and ends with Michelangelo's mastery of it in the first half of the sixteenth No summary of the development of the market for paintings spanning two and a half centuries can do justice to its complexities.^[Interested readers may consult the discussion in @etro2018economics for an economically informed overview of Renaissance art history. For more extensive treatments, see the classic @goldthwaite1993wealth.] However, the following generalizations are meant to  set the stage for the discussion in the next sections.

First, already by the fourteenth century, the market for paintings was regional in scope. While each painter may operate mainly in one city---like Florence and Venice---he accepted jobs from and spent some time in other urban or rural centers. This regional market showed significant degree of competition on both the demand side and the supply side. In the case of the latter, we find a wide array of religious and secular institutions---including guilds and professional associations, monasteries, local governments---as well as private individuals---like wealthy merchants, bishops, and feudal lords, all rivaling each other to secure the services of the most renowned masters of their time. On the supply side, there appears to have been several dozen painters operating in all major art centers. For instance, there were over forty of them in Florence by the fifteenth century [@wackernagel1981world]. Local monopolies or oligopolies were unlikely to emerge and survive since painters were itinerant and guilds seem to have been uninterested or unable to enforce entry restrictions for much of the period [@baxandall1988painting; @thomas1995painter]. Consistent with the above, @etro2018economics finds strong evidence of the law of one price in action in the Renaissance art market.

Second, the two most popular formats for Renaissance paintings were frescoes and altarpieces. The former were larger and generally relied on water-based colors that were applied on walls and ceilings covered in a special plaster while it was still wet. Altarpieces were almost always made of wood panels, also covered in several layers of plaster, but the paint used was egg or oil based. Other formats, though generally less widespread during this period, were banners and canvasses. The latter were especially common in Venice were weather conditions and the proximity of the ocean led frescoes to deteriorate at a faster rate. 

Regardless of the specific format, paintings were complex commodities the production of which required several valuable inputs---including gold, precious stones, wood, and the human capital of the painter. For this reason, a master would generally rely on a number of assistants of varied degree of specialization to aid him in the making of the painting. These formed his workshop and were involved in a variety of tasks, from the production of pigments to the gluing together of wood panels to the drawing and painting of minor figures and backgrounds.^[See the discussions in @thomas1995painter and @piano2021 for more on the division of labor within the workshop.] While the degree of delegation may vary from workshop to workshop, masters were invariably in charge of the design of the painting's composition and major figures, as well as painting all heads and busts.^[We learn from Vasari that at the beginning of the Renaissance, it took one year to complete the average commission. By the sixteenth century, productivity had gone up significantly---likely thanks to the growth in the size of the average workshop---and masters seem to have been able to complete six commissions in one year [@vasari, p. 281].]

Unsurprisingly, paintings were rather expensive. By the second half of the fifteenth century, the price for a panel painted by an established master varied from seven to five hundred Florins for a small piece and a large and elaborate altarpiece respectively [@omalley2013painting, pp. 120-123]. During the same period, a large fresco painting could cost up to three times as much as the most expensive panel painting. For comparison, per capita GDP in the Italian center-north in the year 1500 was approximately 460 Florins.^[Table 1 in @omalley2013painting [p. 120] relies on an implied exchange rate between Florins and Florentine Lire of between 5.6 and 5.8 in the late 1400s. @malanima2011long estimates that per capita GDP in central and northern Italy in 1500 was 81.7 Florentine Lire. @malanima2011long further argues that the silver content of this coin remained constant between 1481 and 1502, so we can estimate that GDP per capita in Florins was approximately 460 Florins.]

In his investigation of the determinants of the price of paintings in Renaissance Italy,  @etro2018economics finds that the value of a commission was a function of a painting's size, number of figures, and the reputation of the commissioned artist. The larger the painting, the higher the price. The same applies to the number of figures. @etro2018economics argues that this may have been correlated with the quality of the painting since masters were less likely to delegate the execution of figures to their less experienced and often less talented assistants. The relationship between reputation and the prices of paintings is more complicated. A good reputation was crucial to succeed in this market [@omalley2013painting]. In deciding whom to award a commission, patrons evaluated a master' skills and creativity by assessing his previous work or by soliciting the opinion of experts, including that of other masters.^[The quality of an artist's past work was, to a considerable extent, public knowledge. Most paintings were meant for public spaces, like churches and municipal buildings, even when the patron was a private individual rather than a corporate entity [@nelson2014patron]. We know from @vasari that Renaissance painters would keep informed on each other's latest work and often traveled to distant towns for the sole purpose of observing or studying a new painting.] Patrons cared also about a painter's reputation for organizational and managerial abilities [@welch2000art, p. 79] as well as timeliness and honesty. Predictably,  @etro2018economics finds that more established masters did earn a premium over beginners. However, his analysis shows that this relationship reverses as artists grow older, possibly due to the weakening of reputational concerns [@omalley2013painting].

Like most commerce in Late Medieval and Renaissance Italy, trade between painters and patrons was governed by a complex system of formal and informal institutions, including professional associations, municipal and religious courts, and private arbitrators [@ogilvie2011institutions; @ascheri2013laws]. While none of the above, taken individually, could do much to prevent opportunistic behavior, their combination contributed to the economic rise of the Italian city-states and they are likely to have facilitated the growth of the market for paintings in the same region as well. Crucial to the effective operation of this complex legal system was the use of written accounts of a deal's terms and conditions. Whenever the commission for a new painting was awarded, patron and master would come together to sign a contract. This was prepared by a professional notary, one of several hundred living in any of the largest Italian cities of the period [@ogilvie2011institutions, p. 294], and its content would be used by private adjudicators---and, if these failed, secular or religious tribunals--to resolve any dispute between the parties involved [@kuehn1987arbitration, p. 299]. The importance of written contracts in Renaissance art markets is confirmed by the fact that---on top of having them certified by notaries, often in the presence of several reputable men acting as witnesses---both parties would often make copies of these documents for their own records, which has allowed some of them to survive until the present day [@omalley2005business].

## The commission contract

A typical commission contract in Renaissance Italy would begin with a religious invocation like "In the name of God amen," followed by the date. If the commission was being awarded by a corporate entity (i.e., a religious order, a guild, or a local government) the document may briefly discuss the deliberative process that had led it to choose a specific artist to undertake the project. The contract would then list the witnesses to its drafting and signing. For each, the contract may note their occupation and place of origin, likely to establish that they were reputable men. Among them was often one notary or legal scholar who may also be referenced as the drafter of the document. The text would then identify the master painter (or painters) who had been awarded the commission, followed by a list of obligations. These may include references to the physical characteristics of the painting---like its size or whether it was a fresco or an altarpiece or some other medium. Often, we find references to the sections the patron expected the painter to perform directly. For instance, a contract may specify that the painter was to complete the painting all by himself or it may say that he was to at least draw and paint the faces and busts of all major figures *sua mano*, by his own hand. Other obligations assumed by the master upon acceptance of the commission pertained to the time and manner of delivery. Thus, a contract may spell out that the painter was to complete the painting of an altarpiece and personally deliver it and install it on the main altar of a given Church within a certain number of months from the signing of the contract. Another item that may be mentioned was that of the master's alternative employment. While most contracts appear to have been silent on the matter, some restricted his ability to accept other commissions until he had satisfied his present obligations.

Frequently, the text of a commission would discuss what the final product ought to look like. It was common to specify what kind of precious materials were to be employed in the production process. For instance, whether the artist was to apply gold leaf on the painting or use pigments made from lapis lazuli, an expensive stone that had to be imported from the Near East. The degree of detail would vary drastically from contract to contract. Some may provide a very precise description of the painting: the specific episode to be depicted---the baptism of Christ, the assumption of the Virgin into heaven, etc.---,which and how many figures where to appear in the foreground and which in the background. Some may mention which colors were to be used to paint a figure's clothing, whether a character was to be standing of kneeling, and so forth. Some contracts, on the other hand, barely mentioned any of these issues, although some may reference that the two parties had agreed to a sketch of the final painting, a copy of which may also be attached to the document.

Repeatedly, commission contracts would use language attempting to set expectations on the quality of the final painting. One popular clause specified that, when gold leaf or pigments derived from precious materials were to be employed, these had to be "fine" or of high quality. Similarly, a contract would often require that the final painting be "at least as beautiful" as some other work by the commissioned master or done with such care that it would be recognized as being of high quality by people with knowledge of such matters.

The text of a commission may mention a variety of financial matters. These include who was to bear production expenses (wholly the painter, wholly the patron, or some combination of the two), the master's compensation, and the method of payment. It wasn't uncommon for a painter to accept a mix of monetary payments---a large payment at the time of the signing of the commission, smaller ones at periodic intervals, and a final sizable handout after completion---and compensation in kind.^[For example, in 1430, painter Antonio di Giovanni d'Anghiari was offered two properties in the town of San Sepolcro in exchange for an altarpiece of his making. See Appendix, Contract #13.] Commission contracts often mentioned penalties both parties were to incur in case they failed to meet their obligations. Thus, a painter may be asked to return some, all, and even twice the money he had received if he did not deliver the painting by the stated deadline. Similarly, the contract may require the patron to compensate the painter for failing to pay him as agreed. Some contracts included a reference to painters' liability over the final product. Such liability may encompass objective and subjective characteristics of the painting. Thus, the commissioning party may demand that the painter fix any damages and imperfections in the painting that may be revealed within some number of months and even years following delivery. The patron may also demand that the artist make a new painting at the latter's expenses if the former is not satisfied by the final product, or reimburse all commission money to the patron.

Finally, a typical commission contract would mention any of a variety of mechanisms for its own enforcement. One popular such mechanism was private adjudication. Patron and painter would agree to have one or more individuals, often art experts or themselves painters. In some cases, the contract may identify the adjudicator or adjudicators by name, other times it would simply state that each party would have the right to nominate an equal number of the committee at a later time. Such a committee was empowered to evaluate the final product and decide whether its value exceeded or fell short of that of the compensation as stated in the commission contract.^[However, the text of the contract would often set limits to a maximum and a minimum value that the committee could not exceed.] Some of the contracts would mention the use of the assets of both parties as collateral. Thus, if either of them failed to meet his half of the deal, the other party may claim ownership of the collateral. Several contracts also noted that liability extended to the family of the signatories. Interestingly, if a painter were to perish before completion of the painting, his heirs would either be required to finish the painting themselves---if they were trained in the craft---or pay for it to be finished by a third party.

# Opportunism in the Renaissance art market

The threat of opportunism is a constant in market interactions.^[Paraphrasing James Madison, @williamson1993opportunism [p. 97] defines opportunism as "the frailty of motive 'which requires a certain degree of circumspection and distrust' [between all parties to a transaction]."] The riches generated by the division of labor and mutually beneficial exchange are, in the presence of transaction costs, bound to attract the ill-intentioned. Renaissance art markets were no exception. Exchange between patrons and masters took place in the context of pre-modern public institutions that were underdeveloped relative to modern standards. Parties could not generally rely on a well-functioning, impartial legal system. This would have been especially true when one of the parties involved (the patron) could be a member of the local ruling class or have close connections to it. Even if they could have relied on effective courts, the nature of the commodity being exchanged (a piece of art) and that of its production process would still have made the problem of enforcement quite challenging. The quality and value of an artistic commodity, especially in the context of a commissioned pieced of art, are ultimately subjective, in the sense that objective variables such as size, the number of figures depicted, the number of hours put in by the painter and so forth are not reliable proxies for quality.^[This point has been noted by @etro2013market [p. 395]. In sixteenth century Venetian art markets, they write: "Contracts were incomplete, however, because the quality of the paintings could be judged by the commissioner (or by dealers and other painters as advisers), but it could be hardly defined ex ante in the contracts.] Reading through accounts of the painting business in Renaissance Italy from primary and secondary sources confirms that the two sides of the art market---but especially patrons, it appears---were concerned about opportunism, and with reason.

## Precious materials

A recurrent object of patrons' concern was opportunism in the handling of all precious materials employed in the production of a painting.^[See @omalley2005business for a comprehensive treatment of this issue.] For much of the Renaissance, patrons insisted to have large sections of the artwork covered in gold.^[This practice went out of fashion by the end of the fifteenth century.] The gold had to be made into thin squares that were then beaten onto the surface of the wall or wood panel before any painting had been done so that a significant portion of the gold would eventually be covered by paint. The amount of gold used would vary from commission to commission, but art historians agree that it represented a significant portion of the total cost of a painting. For instance, @omalley2005business claims that, on average, the cost of gold accounted for a third of the commission price. Relying on data from the workshop of Renaissance painter Neri di Bicci, @diorio2013production comes up with a smaller estimate of about ten percent. Nevertheless, patrons feared that painters would pocket most of the money meant for the purchase of gold leaf and use cheaper substitutes. Indeed, in his *Craftsman's Handbook*, a popular introductory text among aspiring painters of the fifteenth century, Cennino Cennini lamented the many painters had been "embellishing a wall with golden tin, because it is less costly" [@cennini1933craftsman, p. 60].^[Ironically, Cennini rapidly moves on to instruct the reader on how to do just that.] Similar concerned extended to the painter's handling of ultramarine blue. This was a pigment obtained from the grinding of lapis lazuli, a very rare stone, making it the most expensive of colors---fifteen times more expensive than its closest substitute, German Blue [@diorio2013production, p. 50]. Such concerns were not misplaced, as evidenced by the fact that Cennini's *Handbook* contains directions on how to "imitate an ultramarine blue in fresco" involving the mixing of the much cheaper indigo and white, with the addition of small amounts of ultramarine "in the accents" [@cennini1933craftsman].

Though not a precious material, wood was a key input into the execution of commissioned altarpieces. Commission prices would have had to account for expenditures on the wood necessary to make the panels and the, if desired, the frame for the altarpiece. And these were significant expenditures. According to calculations by @diorio2013production [p.40], the amount spent on wood to execute a commission was comparable to that spent on gold and larger altarpieces required more money be spent on it. And so, the same forces that would have led a painter to cheat the patron on the use of gold and ultramarine would have been present here. For example, he could have saved on wood expenditures by simply purchasing a smaller, thinner wood panel from a trusted carpenter.

## Quality, authorship, and subcontracting

Patrons were similarly preoccupied about the possibility of subcontracting. The historical evidence suggests that, in awarding a commission, a patron expected the final painting to be the work of the master receiving it---if not entirely, at least in its most significant aspects, from the design of its final look to the painting of all major figures.^[See @piano2021 for a discussion of the division of labor in Renaissance painters' workshops and masters' involvement in the production of paintings.] Renaissance art scholars have argued that this concern was due to social forces influencing the demand for high art in the Renaissance. Patrons' derived satisfaction from the ability to showcase the work of the most revered painters of their time, since this allowed them to signal their "magnificence" to the rest of society [@nelson2014patron]. A window into the motivation of patrons' demand for painters' work comes from the correspondence between Isabella d'Este---the *de facto* ruler of Mantua in the early sixteenth century---and all major masters of the period. Isabella's letters^[A sample of these letters can be found in @chambers.] show her determination in obtaining paintings from specific artists to add to her collection. Given the growing import of authorship on the value of art, a patron may have feared that, after having received a hefty commission on, a master would subcontract the work to his assistants or some other painter instead. This fear was not without basis. All most prestigious Renaissance painters received a constant flow of requests for their services. Although they would not have been able to complete all the commissions they were offered by themselves, they could have increased their ability to meet demand by delegating much---if not most---of the work to their subordinates. One constraint on excessive delegation came from the threat of reputational losses. Masters would not have wanted to delegate past the point where the quality of the final product was discernibly lower than that of his previous work, or where the hands of his assistants could be easily identified by others. Raphael's execution of the Apostolic Palace in the Vatican is a perfect case study of the dangers of subcontracting. Raphael entrusted the painting of one of the rooms almost entirely to an assistant. The result was, in the words of @shearman1983organization [p. 52], "a catastrophe ... a composite of some half-understood recollections of Raphael’s works." In order to avoid similar scenarios, masters trained their assistants to specifically reproduce their style so as to give the painting a consistent appearance [@comanducci2000fattori, p. 41]. Even though intensive training may have reduced the effect of delegation on quality, it still would have affected the patron negatively if the value of the art to the latter was a function of authorship as well as of quality, as seems to have been the case [@nelson2014patron]. 

Of course, this is not to say that quality did not enter the patron's payoff function. However, as we discuss above, the quality of a painting is a highly subjective matter that cannot be contracted upon easily. To complicate matters farther, Renaissance sensibilities had a complex view of what contributed to the quality of a work of art. Perhaps the first to summarize this view towards the end of the Renaissance, Vasari argued that a painting's quality was a function of an array of variables: Proportion; design; invention; and grace [@vasari, p. 280]. Simply put, a painting was perceived as a masterpiece to the extent that all its figures, objects, and backgrounds had been carefully planned and drawn as realistically as possible, its design and composition were original, and it gave the impression that it had been made effortlessly. Most artists excelled in some but not all these feats, resulting in a unique style that influenced their reputation. Indeed, in a private correspondence between a Milanese patron and his agent in Florence, quoted by @baxandall1988painting [40], the agent expressed the difference between four accomplished masters in terms of their styles. That of Botticelli he characterized as more "virile," that of Filippino Lippi as "sweeter," Perugino's work was "angelic" while Ghirlandaio's was simply "good." In a similar vein, Vasari informs us that the popular Sienese painter Simone Martini "possessed a natural talent for invention" but "did not excel in design" [@vasari, p. 42] and that Raphael's greatest strength was in design and composition [@vasari, p. 332].

An opportunistic master could have exploited the complex nature of artistic quality in any number of ways. He may have put little effort into the design of the scene or in the studies of the main figures. He may have painted a large wood panel or fresco wall with only a few human figures, which required more work, and filled the gaps between them with plain backgrounds. Or he may have drawn the faces and bodies of main figures sloppily or bungled the foreshortening and thus give the painting an awkward and uneven look. And indeed they did, as Vasari mercilessly reports errors and shortcomings from the work of all most established painters of the Renaissance. One egregious case of opportunism in matters of quality was Perugino's practice of employing the same design, down to the same poses and facial expressions for all major figures, for several paintings, what @shearman1983organization [p. 44] has referred to as Perugino's cannibalization of his own work. @omalley2013painting provide extensive visual evidence of this pattern, which was not limited to Perugino. However, the latter may have taken this strategy farther than many of his contemporaries, especially later in his career, leading to such a significant hit to his reputation that he had to move from Florence.^[This infamous episode is recounted by @vasari[p. 265].] The example of Perugino is representative of the fact that reputational concerns could not eliminate the problem of quality opportunism entirely, especially as a painter approached old age. Consistent with this, @etro2018economics finds evidence that commission prices fell significantly in the last few decades of a  master's career, when artists could reap the benefits of a reputation built over their career at little expected cost.

## Late work

Patron-master relationships suffered from more mundane sources of opportunism. Most prominent among them was the propensity of painters to miss deadlines and take longer, often much longer, than agreed upon to complete and deliver the commissioned artwork. Indeed, according to @welch2000art [p. 112], most disputes between patrons and artists were about the latter failing to meet the agreed-upon deadline. For instance, an agent of Isabella d'Este wrote in his response to the latter's inquiry on Pietro Perugino that "he hardly ever finishes a work he has once begun, so long does he take."^[Letter from Francesco Malatesta to Isabella d'Este, September 23rd, 1502, quoted in @chambers[p. 134].] Perugino lived down to his reputation: Having promised to deliver a painting on canvas with a mythological subject, years later the master had not even started to work on it. This prompted Isabella to lament, in a letter to a middleman between her and Perugino, that "we do not know who finds the slowness of these painters more wearisome, we who fail to have our [commissioned painting] finished, or you who have to devise new schemes every day, which then, because of the bizarre ways of these painters, are neither done as soon nor drawn in entirety as we would have wished."^[Letter of Isabella d'Este to Paride Ceresara, November 10th, 1504.] Only after Isabella had threatened to award the commission to another master did Perugino finally bring himself to finish the painting, a full three years after the original deadline. The Augustinian monks of San Donato were not so lucky. They had commissioned Leonardo da Vinci to paint an *Adoration of the Magi* within two and a half years, but the painting had not even started ten years later when da Vinci left Florence for Milan in 1492. The monks waited a few more years, but eventually realized that, if they wanted the painting to be made, they had to find some other painter. Filippino Lippi, the lucky recipient of the new commission, delivered the painting to the monks in 1496.


```{r, echo=FALSE}
# Delayed completion was not always and necessarily the result of opportunism. There was an array of legitimate causes of late work. Events far and away may slow the arrival of materials necessary to the production process like pigments, precious metal, wood, limestone, and so forth. Or the master may realize that the execution of his vision would take longer than originally planned only as he is well into the process. For example, Vasari recounts the difficult relationship between Michelangelo and the Pope Julius II who had hired him during his long stay in Rome. While the Pope were big fans of Michelangelo's work, they found his perfectionism frustrating. Thus, one day Julius II inquired with Michelangelo how much longer it would take him to complete the painting of the Sistine Chapel: "Michelangelo replied that the work would be finished 'when it satisfies me in its artistic details. 'And We' remarked the pope, 'want you to satisfy Us in Our desire to see it done quickly'" [@vasari, p. 442.]. Some time later, when Michelangelo asked him for permission to leave Rome for a brief vacation, the Pope asked him when he would finally be done with the chapel. Michelangelo's response was the same as it had been before, which enraged Julius II. According to @vasari [p. 450], "the Pope, who had a staff in his hand, struck Michelangelo with it as he declared: 'When I can, when I can: I'll make you finish it myself!'." 

# Yet, numerous instances of delayed completion do seem to have been the result of opportunistic behavior. In many cases, this was due painters' tendency to overcommit. Thus, a master may accept a lucrative commission in Florence while he had already committed to completing a painting in nearby Arezzo. Due to the limits in transportation technology, there was no way he could oversee the execution of the two paintings without having to spend many days traveling between the two cities, further slowing down the production of both. Thus, for instance, Perugino's inability to deliver paintings in time was, at least in part, due to the fact that he would accept several commissions at once, often in different cities, forcing him to leave one work unfinished for long stretches of time in order to begin painting somewhere else. Finally, a painter's lateness may be the result of laziness or distraction. For instance, Vasari recounts the troubled history of the commission for a fresco cycle awarded to Raphael by the prominent banker Agostino Chigi:

# >Raphael could not really put his mind to his work because of his love for one of his mistresses; Agostino became so desperate over this that, through his own efforts and with the assistance of others, he worked things out in such a way that he finally managed to bring this woman of Raphael's to come and stay with him on a constant basis in the section of the house where Raphael was working, and that was the reason why the work came to be finished.^[@vasari [p. 328].]
```

## Patron opportunism

Masters were not the only actors capable of acting opportunistically. Patrons also could. Patrons' opportunism could take one of two forms: Failure to compensate painters fully and disingenuous disagreement over the quality of the final painting. Many commissioned paintings were made with the patron's identity in mind. It was common for the altarpiece for a Church dedicated to a specific Saint would often include the same Saint as a main figure---for instance, kneeling to the right of the Virgin Mary holding the child Jesus. Also common was for the patron to demand that his own figure be included in the painting. Moreover, the size and shape of the altarpiece were also tailored to its particular destination. Thus, commissioned paintings took the form of a highly specific investment on the part of the painter and it would have been quite hard to market the artworkk to anyone by the original patron---or at least not at the agreed upon price. The argument applies to frescoes as well, and even more so since the costs of moving the painting to a different location were order of magnitudes higher than for an altarpiece. As it is generally the case with work done on commission, buyers may be tempted to threaten to renege on their promises and offer a lower compensation. A patron could blame this on unexpected financial difficulties or lack of satisfaction with the painter's work, knowing that the latter has very limited options besides accepting a lower compensation.

## Other sources of conflict

Conflict between patrons and painters may emerge in the absence of opportunism. The divergence between patrons' preferences and artists' long-run reputation was one potential source of conflict. For instance, a commissioning party may prefer compositions or subjects that are not aligned with those of most other patrons, present and future, or of other painters and critics. Thus, @welch2000art[p. 114] reports that two Florentine painters had to rebuff the demands of a wealthy merchant for a specific design with the argument that they did not wish to "make fools" out of themselves just to satisfy his unconventional taste. The correspondence between Isabella d'Este and Ventian master Giovanni Bellini is illustrative of this point. The former wanted to add a painting by the latter to her private collection. However, she wasn't merely interested in any painting, but felt very strongly about the specific genre and subject (a classically inspired allegorical fantasy). Unfortunately, Bellini also felt very strongly, just in the opposite direction. He would much prefer to make a more traditional, religiously inspired painting of the Nativity. Indeed, Bellini shared with his peer an aversion to patrons dictating the subject and look of a painting. In a letter to Isabella by her friend Pietro Bembo, who has been acting as middleman between Bellini and her, we read that "[Bellini] does not like to be given many written details which cramp his style; his way of working, as he says, is always to wander at will in his pictures, so that can give satisfaction to himself as well as to the beholder."^[Letter from Pietro Bembo to Isabella d'Este, January 1st, 1506, quoted in @chambers[p. 131].] Eventually, Bellini had the upper hand, although Isabella resumed insisting that he also make an allegorical painting soon after he had delivered the Nativity.

Finally, there was the matter of unexpected death. Renaissance Italy was a dangerous place. The lives of patrons and artists was threatened by a myriad of diseases. The bubonic plague was endemic in Western Europe throughout this period and even minor infections could lead to one's death. There are several recorded instances of masters dying suddenly and at young age. Masaccio, one of the great innovators of early fifteenth century painting, died at twenty-seven years old of unknown causes. Raphael, the great master, died at age thirty-seven, possibly of a sexually transmitted disease. Even older masters would still be in the process of completing a commission at the time of their death. For example, Filippo Lippi died in his early sixties while still working on a fresco cycle for the Cathedral of Spoleto. Similarly, while he lived into his late eighties, the great Venetian painter Giovanni Bellini was still working on a commission awarded to him by the Duke of Ferrara.

# How Renaissance art contracts addressed opportunism

The content of Renaissance art contracts was clearly influenced by the concerns of both parties for opportunistic behavior. For instance, their texts often contained descriptions of what the final painting ought to look like, in an attempt to restrict the artist's ability to cheat by painting fewer figures or less elaborate scenes. In some cases, the contracts may contain just a brief reference to the subject of the painting. For example, the commission between Pietro Perugino and the Servites of Porta Eburnea mentions that the painting was to represent a Madonna with Child Jesus---a popular subject for altarpieces in the Renaissance---but offers no further specifications.^[See Appendix, Contract #72.] In other cases, the commission contract would contain an extremely detailed description of the final painting. An extreme case of the latter is Isabella d'Este commission of a canvas to Pietro Perugino, which includes a long explanation of every element of the painting. Therein, we read that "Pallas should seem almost to have vanquished Cupid, having broken his golden arrow and cast his silver bow underfoot; with one hand she is holding him by the bandage which the blind boy has before his eyes, and with the other she is lifting her lance and about to kill him. By comparison Diana must seem to be having a closer fight with Venus for victory. Venus has been struck by Diana's arrow only on the surface of the body, on her crown and garland, or on a veil she may have around her; and part of Diana's raiment will have been signed by the torch of Venus, but nowhere else will either of them have been wounded."^[Chambers' translation of the original [@chambers].] And on and on for several paragraphs.^[See Appendix, Contract #68.]

@etro2018economics and @etro2013market note that the number of figures in Renaissance and Baroque paintings may have been used as an indirect mechanism to increase artist effort and, thus, the quality of the final artworkk. Since masters were generally in charge of the painting of human figures, by demanding that more figures be included in the painting the patron was ensuring a higher contribution of the master relative to that of his assistants. On their part, masters were aware that patrons and critics would have been better positioned to spot the hand of one of his assistants in the drawing and painting of human figures since these sections were particularly challenging and only talented and experienced painters could generally execute them properly.

Mentions of the number of figures may have also been motivated by the concern that masters may fill the paintings' surface with easy to draw backgrounds and other minor elements. Out of similar concerns, contracts may include explicit references to the size of the final painting. This would have been a more pressing concern for altarpieces, banners, and canvases rather than frescoes, since the size of the latter was always given to the master by that of the wall or ceiling he was to paint. Thus, the commission contract for the painting of a banner from 1466 awarded by the Confraternity of the Annunciation to Piero della Francesca contained the requirement that the banner be "three arms and a half tall and large two and a half arms."^[See Appendix, Contract #34.]

Alongside more indirect mechanisms to encourage artist effort, we find direct references to the parties' expectation of his involvement as well as the quality of the painting. Perhaps the most famous formula introduced by Renaissance art contracts was the *sua mano*---literally translated as "his hand." While the exact meaning of this formula might have evolved slightly during the Renaissance, it was always meant to make explicit the degree of involvement of the master who had been awarded the commission. Thus, some contracts may state simply that the painter was to complete the painting by his own hand, without reference to any specific sections or figures. We find an example of this in Luca Signorelli's commission with the Opera del Duomo di Orvieto in 1500, which reads "and said work master Luca himself is required to execute and paint by his own hand."^[See Appendix, Contract #64.]

Some commissions were more specific. For instance, the contract between Bernardino di Betto (also known as Pinturicchio) and Cardinal Piccolomini in 1502 references the former's obligation "to produce all the drawings and stories on paper and on the wall; to paint all the heads in fresco and make the final touches in secco to his own perfection."^[See Appendix, Contract #67.]

In the attempt to constrain the artist's opportunism, contracts contained explicit references to the quality of the commissioned painting. Such references could take one of four forms: Reference to an absolute level of quality; reference to a level of quality relative to the master's previous work; reference to a level of quality relative to some other painter's work; reference to the consensus of other masters in matters of quality.

We find an example of the first type in a commission contract from 1485 between Domenico Ghirlandaio and Giovanni Tornabuoni.^[See Appendix, Contract #47.] This document instructs Ghirlandaio "to decorate the said chapel with noble, worthy, exquisite and decorative paintings.^[Chambers' translation of the original [@chambers].] An example of the second case appears in the contract for an altarpiece between Benozzo Gozzoli and the Compagnia della Purificazione.^[See Appendix, Contract #30.] Here, we read that "the said Benozzo is obliged to paint it in such a way that it exceeds all the good painting so far done by the said Benozzo, or at least equals it."^[Chambers' translation of the original [@chambers].] A specimen of the third type can be found in the contract between Lorenzo Lotto and Confraternita di Santa Lucia in Jesi. This directs Lorenzo that the painting had to be "better and more beautiful than the altarpiece of the Scuola del Gesù in the same city of Jesi."^[See Appendix, Contract #77.] An instance of the fourth kind of reference to quality appears in the text of a commission between Pietro Perugino and the Arte del Cambio of Perugia, from 1496.^[See Appendix, Contract #59] Here, we read that Perugino agreed to execute a painting that would be perceived as "good and perfect in the opinion of other good and qualified masters and of people with knowledge of this art.

Commission contracts made also frequent reference to the type, quantity, and quality of precious materials the artist was expected to employ int he execution of the commission in an effort to mitigate his temptation to use substitutes of lower quality and price [@omalley2005business]. According to @wackernagel1981world [p. 327], patrons adopted two strategies to prevent artists from cheating on their use of expensive colors and metals. The first strategy consisted of making explicit note of the quantity and quality of these materials in the text of the contract. Thus, it was common practice to make reference to the master's obligation to employ "aurum et azurum finum"---fine gold and blue. 

The second strategy was extra-contractual. The patrons would supply all precious materials themselves instead of paying the master to purchase precious materials, in which case they would have been tempted to buy cheaper alternatives and keep the difference. This strategy has the clear advantage of making it harder for a painter to use substitute cheaper materials for expensive ones. However, it did not make it impossible. For instance, having been handed the latter, a master could sell all or some of it or save it for some other commission. Thus, Vasari reports a humorous episode from the life of Pietro Perugino in which the patron (the prior of some religious order) had supplied Perugino with "a great abundance" of ultramarine blue "[and] wanted Pietro to employ it lavishly." However, "[he] was so miserly and suspicious that he did not trust Pietro and always wanted to be present when Pietro used this color in his work" [@vasari, p. 260]. 

We observe a similar strategy adopted for the commissioning of altarpieces that required the use of wood for panels and frames. A patron may supply panels and frames directly to the painter, thus reducing the latter's ability to employ cheaper wood for both. This strategy had the further advantage of leaving little room for the painter to produce a smaller painting. Alternatively, the two parties would specify the size of the final artworkk in the text of the contract.

Enforcing deadlines on artists seems to have been exceptionally challenging. Not even the Pope could, as a patron to Michelangelo, do much to expedite the process. Thus, commissioning parties had to come up with clever solutions to the problem of overcommitted and/or lazy painters. One such solution was to include in the text of the contract explicit restrictions on alternative sources of employment until the master had completed and delivered the present commission. Thus, we read in a contract between Bartolomeo Montagna and the Scuola Grande di San Marco in Venice that the former accepted "make the aforementioned painting and not work on anything else."^[See Appendix, Contract #43.] In a different contract, we learn that Benedetto Bonfigli---awarded a commission by the City of Perugia---agreed "to work continuously on painting in the aforementioned chapel whenever he can. And, for the duration of such work, he may not accept any other work until he has finished and delivered the present one."^[See Appendix, Contract #35.]

A more straightforward method to prevent delays was to include monetary penalties for late work. For example, in 1404, Taddeo di Bartolo agreed to complete a fresco for the Duomo of Siena within one year and to forego the final installment of his compensation if he failed to meet this deadline.^[See Appendix, Contract #7.] Similarly, in a contract from 1466, Benozzo Gozzoli agreed to pay the city of San Ginignano twice the amount of the commission in case he failed to deliver an altarpiece within four months.^[See Appendix, Contract #33.]

Keenly aware of the potential for conflict between them, the parties consenting to the terms of the commission contract would often also agree to subject themselves to a dispute resolution mechanism. A popular such mechanism was the *lodo*. In the case of disagreement between patron and master over the quality of the final painting, commission contracts may assign the task of adjudication to a committee of two or more experts---often local painters---to be selected in equal number by either party. For example, Priamo di Pietro della Quercia the representatives of the Church of San Michele in Volterra agreed that any disagreement on the value of the painting was to be adjudicated "by two citizens of Volterra, one elected by [the Church of Volterra] and the other by said master Priamo."^[See Appendix, Contract #17.] In some other cases, instead of a committee, the contract would identify a sole individual who was recognized by both parties as competent in such matters. Thus, Domenico Ghirlandaio and patron Giovanni Tornabuoni agreed, in a contract from 1485, to have the final work evaluated by "a just man."^[See Appendix, Contract #47.]

While the *lodo* may help resolve disputes about the quality of the artworkk, it could little to address the many other sources of opportunism. Hence, Renaissance art contracts relied on two further instruments to ensure compliance. The first was the use of delayed payment. Instead of transferring the entire compensation amount to the painter at the moment he was awarded the commission, the contracts would specify that payment would be made in several installments for the duration of the commission and that the final installment would be paid only upon completion and delivery. For instance, the contract between Pietro Perugino and the Arte del Cambio di Perugia from 1496 includes the clause that Perugino was to receive "fifty ducats presently at the start of his work, and fifty ducats upon completion of the fresco, then fifty more at the beginning of his work on the altarpiece, and from then on fifty ducats a year each year until he has received the three hundred and fifty ducats he is owed."^[See Appendix, Contract #59.]

The second tool for mitigating opportunism we observe in Renaissance art contracts is the use of collateral. Several commissions recorded the master's agreement to collateralize his assets to guarantee his performance. For example, we read in Lorenzo Lotto's contract with the Compagnia of Santa Lucia in Jesi that the master had set as collateral "all assets present and future."^[See Appendix, Contract #90.] The same text contains a reference to the fact that patrons had agreed to the same. Moreover, we read that each party was entitled to "induce, force, compel, motivate, banish, tie, imprison [the other party] until all obligations have been met." This is indicative of the fact that, in drafting Renaissance art contracts, both sides of the market understood the potential of patrons' opportunism.   

Finally, commission contracts made explicit reference to the possibility of the death of the master before the painting had been delivered. In such an eventuality, his heirs would have been required to return all the money he had been paid to execute the commission. Thus, the text of the commission between Perugino and Isabella d'Este reads "that in the event of the death of the said Master Pietro, should it happen that the said work is not completed, the heirs of the said Master Pietro shall be obliged to restore the said sum of 20 large gold florins to [Isabella's agent]."^[Chambers' translation of the original [@chambers]. See Appendix, Contract #68.]

# Understanding Renaissance art contracts

## Renaissance art contracts as data

To investigate the role of formal contractual obligations in the mitigation of opportunism in the art market of Renaissance Italy, we created a new data set of commission contracts from this period. We started with the list of contracts mentioned in @omalley2005business and tracked down all the sources for the transcriptions of these texts. We then consulted a variety of secondary sources, including monographs on specific Renaissance painters, schools, and periods.^[A full list of sources for Renaissance art contracts can be found in **APPENDIX**.] Finally, we searched the database of documents from this period at the *Italian Renaissance Document Site* for the transcription of contracts between patrons and painters.^[The database can be accessed at http://irds-project.org/.] 
All in all, our final data set consists of ninety-one documents, each corresponding to an individual commission. Seventy-seven of these are the texts---in many cases, partial texts---of the commission contracts themselves; the remaining fourteen documents are summaries of the contracts, made contemporaneously by either the patron or the painter for accounting purposes.

We analyzed each of these documents to extract document-level variables about their structure and content.^[We describe our method in detail in **APPENDIX**.] Informed by our discussion of opportunism in Renaissance art markets, we asked whether each text references any of ten broad categories of contractual terms and conditions. Table 1 lists the ten categories and provides a brief description of each. We used this information to create a set of indicator variables associated with these categories and assigned a unit value to the variable if the text of the documents references explicitly one of our items---for instance, the use of precious materials or the expected date of delivery of the painting---and a value of zero otherwise. Table 2 provides a summary of our findings. For each item, the table shows the percentage of documents in our database that contain the same item

**Table 1: Categories of contractual terms and conditions**

|Category|Description|
|:-|:-:|
|Look|*Does the commission document mention the subject of the painting or its size?*|
|Materials|*Does the commission document specify any of the materials to be used in the making of the painting?*|
|Quality|*Does the commission document reference the absolute or relative quality of the painting?*|
|Deadline|*Does the commission document specify a delivery date?*|
|Compensation|*Does the commission document contain the total compensation to the painter?*|
|Enforcement|*Does the commission document make reference to specific enforcement mechanisms?*|
|Authorship|*Does the commission document contain specific restrictions on the artist's ability to subcontract?*|
|Delivery|*Does the commission document specify how the contract is to be delivered?*|
|Employment|*Does the commission document contain restrictions on the artist's ability to accept other jobs?*|
|Expenses|*Does the commission document specify how expenses were to be shared between the agreeing parties?*|

Unsurprisingly, the most mentioned category is that which pertains to the compensation of the painter. Over ninety percent of all documents in our data base contain at least one reference to the amount the patron was expected to pay to the commissioned artist. Two other common categories mentioned in our documents are those pertaining the look---the subject of, colors to be used in, and the size of---the painting, with almost ninety percent, and enforcement---any mention of how contractual violations and other kinds of conflict between the trading partners were to be resolved---, with just short of eighty-seven percent. The least recurring category, on the other hand, is delivery---whether the contract specifies how the final painting was to be transported to its ultimate destination---^[This is not surprising since a significant portion of our commission documents refer to the production of frescoes, which need not be transported to their final destination.] with fourteen percent, followed by the category employment, which captures whether the contract-imposed restrictions on the painter's freedom to accept other jobs for the duration of the commission, with just shy of nineteen percent.

**Table 2: Summary statistics, dependent variables**

|Variable|N|Mean|Standard Deviation|Min|Max|
|:-:|:-:|:-:|:-:|:-:|:-:|
|Look|91|0.89|-|0|1|
|Materials|91|0.659|-|0|1|
|Quality|91|0.659|-|0|1|
|Deadline|91|0.681|-|0|1|
|Compensation|91|0.912|-|0|1|
|Enforcement|91|0.824|-|0|1|
|Authorship|91|0.253|-|0|1|
|Delivery|91|0.143|-|0|1|
|Employment|91|0.187|-|0|1|
|Expenses|91|0.67|-|0|1|
|Index|91|5.879|1.467|1|8|
|Length|91|476.824|345.368|101|2,176|

We use this information to generate an index of contractual completeness. For each document, the index measures how many of these ten items or categories it references explicitly. The average document in our database has an index value of 5.88. The lowest value of the index in our data set is 1, the highest is 8, meaning that none of the documents mentions every one of our categories. Figure 1 shows the distribution of the values of our index across the ninety-one commission documents. Finally, the table shows that the average commission document was about four-hundred and eighty word-long. This is over four times the length of the shortest text while the longest commission document in our database is over two-thousand word-long. 

**Figure 1: Contractual completeness index across commission documents**
```{r, echo=FALSE, message=FALSE}
library(readxl)
library(stargazer)
library(sandwich)
library(AER)
library(lmtest)
library(tidyverse)
library(haven)
library(plm)
library(ggplot2)
library(rio)
library(openxlsx)
library(dplyr)

Data <- read.xlsx("Data.xlsx")

ggplot(data.frame(Data), aes(x=Index))+
  geom_bar(color="black", fill="grey")+labs(x="Contractual completeness index value", y="Count")
```

## The economics of Renaissance art contracts

Economic theory provides a framework to understand variation in the content and structure of Renaissance art contracts. The key intuition behind the economic analysis of contracts is that drafting the latter---specifying all the trading partners' rights and obligations under all possible future circumstances---is a costly endeavor. As with any costly activity, economics predicts that people will stop as soon as its (positive and increasing) marginal cost approaches its marginal benefit. In the drafting of a contract then, the equilibrium degree of completeness will be determined by the behavior of the marginal benefit of increasing completeness and its marginal cost. Thus, all things equal, an increase in the marginal benefit of increasing the degree of completeness, the more motivated trading parties will be in incurring the extra costs of drafting a more exhaustive contract. This leads us to our first proposition:

*Proposition 1: The more valuable the commission, the higher the degree of contractual completeness*

This approach also emphasizes that people are willing to substitute away from costly activities in cheaper substitutes are available. In the context of the drafting of formal contracts such a substitute is offered by informal or "relational" contracting [@michler2020relational]. Relational contracts rely on implicit and often unwritten agreements and have the property of being self-enforcing: Trading partners comply with the terms of conditions of the agreement out of self-interest without the threat of punishment by a third party, usually the state. While they may be cheaper than formal contracts, relational agreements are rather fragile. In order to work effectively, they require that the parties involved fear large-enough losses if they act opportunistically. For instance, they may fear the loss of future business due to a compromised reputation for honesty [@klein1981role]. Informal contracts will also be more effective when all parties have access to reliable information about their exchange partners. In the case of Renaissance art, information about past performance will be particularly important since the patron is purchasing a commodity before its production has even begun and the painting's final look can never be perfectly specified ex-ante. Which brings us to our second proposition:

*Proposition 2: Contracts for commissions awarded to better-established artists will show lower degrees of completeness*

An artist's reputation will serve the double function of provide patrons with information about past performance and signal that he has something to lose----his hard-earned reputational capital---from acting opportunistically. Established painters would have met both conditions necessary to make relational agreements a feasible alternative to formal contract. Their reputation may work effectively as a substitute for contractual completeness: In hiring reputable masters, patrons may feel less of a need to specify all possible terms and conditions in the text of their commissions.


```{r, echo=FALSE}
# The key intuition of the economic approach to the study of contracts is that drafting a complete contract---an agreement that specifies the consenting parties' rights and obligations under all possible circumstances---is costly. From this starting point, we can develop a framework to explain observe variations in the observed structure and content of business agreements in general and Renaissance art contracts in particular. Since increasing the degree of completeness of any such document is costly, parties will prioritize the inclusion of terms and conditions that---given the characteristics of the actors involved, those of the commodity being exchanged, the physical, economic, and legal environment, the information available to each party, and so on---they expect will yield the highest net benefits.
# Combined with institutional and historical knowledge of the art market in Renaissance Italy, this framework generates several testable implications about variation between commission contracts from this period.
# *Proposition 1: The more valuable the commission, the higher the degree of contractual completeness*
# All other things equal, parties to a contract will be more willing to invest valuable resources into the costly contracting process the higher the expected returns from such an investment. Thus, we expect the degree of specificity of a commission contract to increase in the monetary compensation offered to the painter.
# *Proposition 2: Contracts for commissions awarded to better-established artists will show lower degrees of completeness*
# Economic agents invest in the accumulation of reputational capital because they expect the latter to facilitate future transactions. Reputation works as a kind of hostage capital the value of which is damaged by one's opportunism. Knowing this, other economic agents will be more trusting of one's promises. Thus, reputation may work effectively as a substitute for contractual completeness: In hiring established, reputable masters, patrons may feel less of a need to specify all possible terms and conditions in the text of their commissions.
# *Proposition 3: The relationship between contract completeness and age of the artist will be U-shaped*
# Reputation will stop functioning as hostage capital as one's time horizon gets shorter. A painter may be less likely to act opportunistically if doing so may endanger decades of fruitful work than one or two years of the same. Knowing this, patrons will tend to be more skeptical of reputable painters the older the latter get. Thus, reputation will be a less effective substitute for contractual completeness when dealing with old masters.
# *Proposition 4: Contracts for more valuable paintings will be more likely to specify enforcement mechanisms*
# *Proposition 5: Contracts for commissions awarded to better-established artists will be less likely to specify enforcement mechanisms*
# *Proposition 6: The probability of a contract specifying enforcement mechanisms will decrease and then increase in the artist's age*
# Setting up enforcement mechanisms can be particularly tricky since parties must detail the circumstances under which such mechanisms must be activated and what rules will govern them. However, the contracting parties will have more of an incentive to do so if they stakes are higher, that is, if the value of the exchange---the commissioned painting---is valuable enough. Similarly to proposition 2, the reputation of the artist may act as a substitute for the need to specify enforcement mechanisms. Thus, contracts for commissions awarded to reputable masters should be more likely to include mentions of the latter once we control for the value of the painting. However, as we discuss above, the effect of reputation is weakened as artists grow older. Hence, we predict that very young and very old masters, even reputable ones, will be more likely to be bound by specific enforcement mechanisms than masters at the height of their careers.
# *Proposition 7: The better-established the commissioned master, the more likely that a contract will include restrictions on alternative employment*
# One drawback of hiring an established master was that, being in high demand, he would be tempted to accepts several commissions at once, which will likely result in delays in the execution of the painting. When drafting a contract, the patron is more likely to insist on including such a provision when dealing with an established artist than with a younger, unproven painter.
# *Proposition 8: The better-established the commissioned master, the more likely that an agreement will include restrictions on subcontracting*
# In order to handle the large workload of profitable commissions they were awarded, established masters relied on the assistance of artisans and artists working within their workshops. The more tasks necessary to the execution of a commission he entrusted on members of his workshop, the more paintings he could take on at any point in time. Aware of this possibility, patrons would have wanted to ensure some minimum level of the master's direct involvement in the execution of a commission. Since younger artists were less likely to operate a large workshop, there would have been less of a need to include such a conditions when dealing with them.
```


## Data

To test our two propositions, we need measures of 1) the value of a commission and 2) the reputation of the commissioned artist. For the former, we extracted the compensation amount mentioned in each commission and used the *Index of the Real Value of the Current Florin of Account, 1300–1600* from @goldthwaite2009economy to express all commission number in 1300 Florins.^[One of the contracts in our database was signed in 1285, fifteen years before the beginning of the data in @goldthwaite2009economy. In our analysis, we assume that no inflation happened in the last fifteen years of the thirteenth century.] The highest paid commissions in our database are from two esteemed members of the Florentine school, Pietro Perugino and Domenico Ghirlandaio. In 1485, the latter was promised eight-hundred and three Florins for a fresco painting. In 1502, Perugino was promised eight-hundred and forty Florins, also to make a fresco. Bartolomeo della Gatta, a Florentine painter himself, has the dubious distinction of being on the receiving end of the two smallest compensations in our data set---from 1475 and 1488---both of which are of just above four Florins. There is no straightforward measure of reputation for Renaissance painters. Thus, we adopt three strategies to proxy an artist's reputational capital. The first strategy is to use age at the time of the signing of a commission. The idea is that a painter's reputational capital will generally increase with age. We rely on a wide array of secondary sources to establish a painter's year of birth and subtract it from the year of the commission.^[All sources we use to come up with this measure are listed in the **APPENDIX**.] Our second strategy is to use the number of years of activity at the time of the commission. We follow a similar process as for age by consulting secondary sources to find the year of the first painting attributed to that artist. We then subtract this from the year of the commission.^[We provide a list of all sources we rely on for this measure in the **APPENDIX**.] and consulted secondary sources for a comprehensive list of all paintings attributed to an artist over his lifetime. Finally, we counted the paintings completed until the year before the signing of the commission. The logic of these latter two measures is that painters with longer careers and more commissions under their belt will generally be better established and enjoy a higher reputation. Following @etro2018economics, we also measured reputation by counting the number of words dedicated by Vasari---the sixteenth century painter and art historian---to each of the artists in our data set in his *Lives of the Artists* [@vasari]. To control for potential confounding variables, we collected information about the type of painting commissioned---whether it was an altarpiece, a fresco, or some other kind---and the identity of the patron---whether it was a religious institution or not. Table 2 provides summary statistics for our independent and control variables.

```{r, echo=FALSE}
# Length in Vasari
# Size
# Distance between patron and painter's workshop
```

**Table 3: Summary statistics, independent and control variables**

|Variable|N|Mean|Standard Deviation|Min|Max|
|:-:|:-:|:-:|:-:|:-:|:-:|
|Paintings|89|21.466|24.551|0|101|
|Active|90|19.161|13.467|0|50|
|Age|85|46.859|14.443|22|85|
|Price|81|98.067|141.142|0.755|840|
|Contract|91|0.846|-|0|1|
|Full|91|0.769|-|0|1|
|Latin|91|0.549|-|0|1|
|Vulgar|91|0.33|-|0|1|
|English|91|0.11|-|0|1|
|Team|91|0.077|-|0|1|
|Artists|91|1.099|0.396|1|4|
|Clerical|91|0.593|-|0|1|
|Altarpiece|91|0.56|-|0|1|
|Fresco|91|319|-|0|1|
|Other|91|0.121|-|0|1|
|Vasari|91|2,549|2,826.258|0|17,300|

## Empirical strategy and results

Our empirical strategy is twofold. First, we estimate two sets of OLS specifications on the effect of commission- and painter-specific characteristics on a document's contractual completeness index and the same document's length---i.e., its word count. Second, we estimate logit regressions for four key categories of contractual terms and conditions: Whether a commission document mentions explicitly the quality of the final product ("Quality") or any element of the painting's look ("Look"), whether it specifies the degree of involvement of the awarded artist ("Authorship"), and whether it details restrictions on the latter's ability to accept other jobs ("Employment").

Table 4 reports the results of the first set of OLS specifications. Columns (1)-(3) include only our variables of interests: The (natural logarithm of) the price of the commission in 1300 CE florins and each of our three alternative measures of a painter's experience at the signing of the contract---his age (column (1)), how long he had been active as a painter (column (2)), and the number of paintings he had completed up to the year before the commission (column (3)). Columns (4)-(6) include a series of control variables, including whether the document was an actual contract or a summary of the terms of the agreement made by one of the parties for accounting purposes, whether the original text was in Latin or Vulgar, whether the commission was for a fresco painting, an altarpiece, or some other medium, and so forth. The results from table 4 show a strong and robust positive effect of the (natural logarithm of) price on a commission document's value in our index of contractual completeness. A one percent increase in the commission price predicts an additional (approximately) 0.004 value in our index of contractual completeness. Recall that commission prices in our data set varied drastically, from less than one florin to eight-hundred florins. In this context, a doubling of the commission price would increase the value of the index of contractual completeness by almost half a point. While none of the coefficients for out three measures of painter reputation are statistically significant, they show remarkable consistency in their sign across all six specifications: Commissions awarded to older painters with longer careers, and more commissions under their belt seem to have been characterized by lower degrees of contractual completeness. The coefficient for the quadratic terms associated with these same variables are similarly consistent in their sign, though they too are not statistically significant and their sign is positive throughout. These results may indicate that reputation did function as a substitute for contractual completeness, but less so for aging painters---a result consistent with the finding in @etro2018economics of painter receiving smaller compensations in the last years of their lives.

**Table 4: **
```{r, echo=FALSE, message=FALSE}
library(readxl)
library(stargazer)
library(sandwich)
library(AER)
library(lmtest)
library(tidyverse)
library(haven)
library(plm)
library(ggplot2)
library(rio)
library(openxlsx)
library(dplyr)


Data <- read.xlsx("Data.xlsx")

# OLS regressions
  # Index

index_1 <- lm(Index~log(Price)+Age+Age2, data=Data)
index_2 <- lm(Index~log(Price)+Active+Active2, data=Data)
index_3 <- lm(Index~log(Price)+Paintings+Paintings2, data = Data)
index_4 <- lm(Index~log(Price)+Age+Age2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
index_5 <- lm(Index~log(Price)+Active+Active2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
index_6 <- lm(Index~log(Price)+Paintings+Paintings2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data = Data)

stargazer(index_1, index_2, index_3, index_4, index_5, index_6, type = "text")
```

Table 5 shows the results of our second set of OLS estimates. Once again, we find a robust, positive, and highly significant effect of price on contractual completeness---in this case the length of the commission document. A one percent increase in the compensation for a painting is associated with at least one extra word increase in the document. A one standard deviation increase in price predicts two-thirds of a standard deviation increase in the length of the commission document. As in table 4, none of the coefficients for our measures of painter experience are statistically significant. Also as in table 4, we find consistency in the sign of their effect. However, this time the effect is positive: Artists with longer careers were awarded commission contracts that tended to be shorter.

**Table 5:**
```{r, echo=FALSE, message=FALSE}
# OLS regressions
  # Length

length_1 <- lm(Length~log(Price)+Age+Age2, data=Data)
length_2 <- lm(Length~log(Price)+Active+Active2, data=Data)
length_3 <- lm(Length~log(Price)+Paintings+Paintings2, data = Data)
length_4 <- lm(Length~log(Price)+Age+Age2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
length_5 <- lm(Length~log(Price)+Active+Active2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
length_6 <- lm(Length~log(Price)+Paintings+Paintings2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data = Data)

stargazer(length_1, length_2, length_3, length_4, length_5, length_6, type = "text")

```

Tables 6, 7, 8, and 9 summarize the results of our four logit specifications estimating the effect of (the natural logarithm of) price and artist-specific characteristics on the probability that a commission document will mention the quality of the painting, and its look, or specify restrictions on subcontracting and on alternative sources of employment for the duration of the job. We find particularly strong evidence of a positive effect of price on whether a document makes explicit reference to the quality of the final painting:A one standard deviation increase in the natural logarithm of the price of the commission increases the odds that a commission document mentions the quality of the painting explicitly by over eighty percent. The predicted effect of price on variables "Look" and "Employment" is never statistically significant. However, we find some evidence of a positive effect on the odds that the text associated with a specific commission will contain restrictions on an artist's ability to subcontract relevant parts of his work---"Authorship." Table 6 also shows that, if we measure experience by the number of paintings completed by an artist until the year before the commission---columns (3) and (6)---an increase in painter's experience is associated with a statistically significant fall in the odds that a commission document will mention the quality of the final painting. None of the coefficients of our measures of painter experience is statistically significant in any of the other specifications in tables 6 through 9. On the other hand, their sign is consistently negative. Across twenty-four specifications and all three measures of experience, the coefficient is negative in twenty occasions and positive in only four. While not conclusive, this is suggestive evidence that experience may have worked as a substitute for contractual completeness in Renaissance art markets. Moreover, in eighteen out of the twenty instances in which we find a negative coefficient on our measure of painter experience, the corresponding quadratic term is positive eighteen times. Once again, these coefficients are not statistically significant, but the consistency of the direction of the effect indicates the possibility of the existence of a U-shaped relationship between reputation and contractual completeness, suggesting that the substitution effect between the two weakened towards the end of one's career and consistent with the findings in @etro2018economics on commission prices and the qualitative evidence discussed in @omalley2013painting.

**Table 6:**
```{r, echo=FALSE, message=FALSE}
# Logit regressions
  # Quality
quality_1 <- glm(Quality~log(Price)+Age+Age2, data=Data)
quality_2 <- glm(Quality~log(Price)+Active+Active2, data=Data)
quality_3 <- glm(Quality~log(Price)+Paintings+Paintings2, data = Data)
quality_4 <- glm(Quality~log(Price)+Age+Age2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
quality_5 <- glm(Quality~log(Price)+Active+Active2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
quality_6 <- glm(Quality~log(Price)+Paintings+Paintings2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data = Data)

stargazer(quality_1, quality_2, quality_3, quality_4, quality_5, quality_6, type = "text")

```

**Table 7: **
```{r,echo=FALSE, message=FALSE}
# Logit regressions
  # Look

look_1 <- glm(Look~log(Price)+Age+Age2, data=Data)
look_2 <- glm(Look~log(Price)+Active+Active2, data=Data)
look_3 <- glm(Look~log(Price)+Paintings+Paintings2, data = Data)
look_4 <- glm(Look~log(Price)+Age+Age2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
look_5 <- glm(Look~log(Price)+Active+Active2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
look_6 <- glm(Look~log(Price)+Paintings+Paintings2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data = Data)

stargazer(look_1, look_2, look_3, look_4, look_5, look_6, type = "text")

```

**Table 8: **
```{r,echo=FALSE, message=FALSE}
# Logit regressions
  # Authorship

author_1 <- glm(Authorship~log(Price)+Age+Age2, data=Data)
author_2 <- glm(Authorship~log(Price)+Active+Active2, data=Data)
author_3 <- glm(Authorship~log(Price)+Paintings+Paintings2, data = Data)
author_4 <- glm(Authorship~log(Price)+Age+Age2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
author_5 <- glm(Authorship~log(Price)+Active+Active2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
author_6 <- glm(Authorship~log(Price)+Paintings+Paintings2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data = Data)

stargazer(author_1, author_2, author_3, author_4, author_5, author_6, type = "text")

```

**Table 9: **
```{r,echo=FALSE, message=FALSE}
# Logit regressions
  # Employment

employ_1 <- glm(Employment~log(Price)+Age+Age2, data=Data)
employ_2 <- glm(Employment~log(Price)+Active+Active2, data=Data)
employ_3 <- glm(Employment~log(Price)+Paintings+Paintings2, data = Data)
employ_4 <- glm(Employment~log(Price)+Age+Age2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
employ_5 <- glm(Employment~log(Price)+Active+Active2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data=Data)
employ_6 <- glm(Employment~log(Price)+Paintings+Paintings2+Contract+Full+Latin+Vulgar+English+Team+Clerical+Altarpiece+Fresco+Vasari, data = Data)

stargazer(employ_1, employ_2, employ_3, employ_4, employ_5, employ_6, type = "text")

```





```{r, echo=FALSE, message=FALSE}
# Table 6 provides the estimates for our two OLS regressions. Columns (1) through (3) show the effect of painter-, patron-, and commission-specific characteristics on the value of our index of contractual completeness. Columns (4)-(6) do the same, but using the length of each document to measure contractual completeness.

# Combined, the results from tables 3-6 allows us to evaluate th empirical validity of our eight propositions. According to proposition 1, the value of a commission should have a positive effect of the degree of contractual completeness of the associated documents. This is arguably the proposition that finds that strongest and most robust confirmation in our data. The coefficient on *Price* has a positive and statistically significant effect on the probability that a document will reference the quality of the final painting---table 3, columns (4)-(6)---, on the value of a document's contractual completeness index value---table 6, columns (1)-(3)---and on the length of the commission contract. The last result is particularly robust, as it is significant at the one-percent level, and is of large economic significance as well: Each extra florin worth of commission predicts over one extra word in the corresponding contract. The evidence is much weaker when it comes to the effect of *Price* on the probability that a commission document will contain a reference to a specific enforcement mechanism, as predicted by proposition 4.

# Consistent with proposition 2, we find evidence that commission awarded to better established painters show lower degrees of contractual completeness. For instance, if we measure the experience of a painter by the number of works completed by the year before the present commission---*Paintings*---we find that this is associated with a lower probability that the associated document will reference the quality of the final painting explicitly---table **?**, column (4), even though the effect is not statistically significant and the sign of the coefficients is inconsistent when we rely on other measures of experience, such as age and years of activity at the time of the commission. Also not significant, but consistently negative across our three measures of experience, is the effect on the probability that the document will specify any aspect of the look of the final painting---table 3, columns (1)-(3). Similarly, the coefficients are negative for two of our three measures of painter experience when predicting the probability that the contract will mention the compensation amount---table 4, columns (2) and (3)---though these too are not statistically significant. Finally, the coefficients for *Paintings*, *Active*, and *Age* are all negative---though not statistically significant---when we estimate the OLS regression for the value of the index of contractual completeness---table 6, columns (1)-(3). According to proposition 5, the same three measures should have negative coefficients when predicting their effect on the probability that a commission document will mention an enforcement mechanism for the agreed-upon terms. The coefficients do indeed have the predicted signs, though they fail to meet the standard thresholds for statistical significance. Two more propositions focus on the effect of a master's experience. Proposition 7 predicts that better-established masters will be more likely to be bound by restrictions on alternative employment. Proposition 8 states that the same masters will be more likely to face restrictions on subcontracting. The results in table 5 provide little evidence in support of either proposition.

# Finally, propositions 3 and 6 predicted a U-shaped relationship between a painter's experience and both the degree of a commission's contractual completeness and the probability that the commission document will mention an enforcement mechanism explicitly. We find evidence consistent with the former. For instance, for all three measures of artist experience including the quadratic term has the opposite (i.e., negative) sign of the non-quadratic term---table 3, columns (1)-(3). Wehn we estimate the coefficient of the effect of the quadratic term for *Paintings* on *Quality*---table 3, column (4)---, we also find a negative and statistically signficant---at the five-percent level---value, also consistent with a U-shaped relationship between artist experience and contract completeness. Two out of three coefficients for the quadratic terms of our measures of artist experience are positive while those of the non-quadratic terms are negative when we estimate their effect on *Compensation*---table 4, columns (2) and (3)---though not statistically significant. Also not statistically signficant but consistent across all three measures, we find a U-shaped relationship between our three measures and our index variable for contractual completeness---table 6, columns (1)-(3). We do not find any such regularity when we estimate the effect of the same variables on the length of the commission documents. Nor do we find strong evidence for the prediction in proposition 6. Our quadratic terms---table 4, columns (4)-(6)---are never statistically signficant. However, in two out of three instances, the signs are consistent with a U-shaped relationnship with *Enforcement*.

```


# Conclusion

Renaissance paintings were complex commodities exchanged between parties---the patron and the master---who operated in a world of uncertainty---about the quality of the final painting, the possibility of patron bankruptcy, the potential death of the artist before the painting was completed---and of asymmetric information---about the effort of the painter, the quality of the material, the involvement of assistants. Moreover, the opportunity cost of most of the inputs involved---the human capital of the artist, the wood and precious materials used in the execution of the painting---were quite valuable, meaning that the price of the benefits from the exchange to both parties was high, but so were those from opportunistic behavior. In this paper, we study the nature and consequences of opportunism in the production and exchange of Renaissance paintings. We show that the nature of the execution of these commodities was potentially conducive to cheating. Painters could substitute cheap materials for expensive ones, they delegate significant portions of their responsibilities to less skilled artisans, and may take much longer than promised to deliver the final product. Patrons could also act opportunistically, for instance, by lying about their satisfaction with the complete painting in order to pay a lower price for it. The exchange was threatened by non-opportunistic considerations as well. Patron and painter may genuinely disagree about the aesthetic value of the final product, since taste in matters of art are ultimately subjective. Or, either party may die suddenly, a possibility that was unfortunately quite common in Renaissance Italy.

Building on insights from the economic analysis of contracts, we identify a set of propositions about the impact of the threat of opportunism on the contracting process between patrons and painters. Using a novel data set of ninety-one commissions from Renaissance Italy, we find that painter-, patron-, and artwork-specific characteristics influenced the content of these documents in ways consistent with the predictions of the economics of contracts. Specifically, we find evidence that the texts of more valuable commissions presented a higher degree of contractual completeness. We also find suggestive evidence that a painter's reputation worked as a substitute for the writing of formal contracts. Both results are consistent with key insights from the economic approach to the study of contracts. 





# REFERENCES











